Modelling the City of Melbourne Pedestrian Data

### 2017-09-30 presentation/final.Rmd
update

We use the daily data sourced from `pedestrian.melbourne.vic.gov.au` rather than the data available at the Open Data Platform as missing values are explicit

@@

```{r model_alg, echo = FALSE, warning=FALSE, message=FALSE, cache = TRUE}
n_obs <- nrow(dfa)
threshold <- 0.1
threshold_miss_hrs <- 50

@@

At each sensor location with small proportion of missing values, we estimate a generalised linear model with:
$$\mu_{\text{Sensor}, \text{Month}, \text{DayType}, \text{Time}} \sim \text{Month} + \text{Time}\times \text{DayType}$$
and a quaispoisson error distribution.
- Quasipoisson regression allows for $Var[\text{HourlyCount}] = \theta \mu$
- These models only use time and date based variables to predict pedestrian counts
  * `DayType` is a factor level/categorical variable based on the day of the week and public holidays
  * `Time` is treated as a factor level, as it has a non-linear relationship with counts

@@

```
- Need to use complete data from neighbouring sensors
  * Use data from imputation in previous step
- Neighbour defined as geographical neighbours for simplicity
  * Find two closest sensors by great-circle (haversine) distance

@@

---
## _Step 2c: Fit models using neighbours_
At each sensor location with "large" proportion of missing values, we estimate a generalised linear model with:

@@

>>>>

We use the daily data sourced from `pedestrian.melbourne.vic.gov.au` rather than the data available at the Open Data Platform as missing values are explicit
- accessible using `rwalkr::walk_melb()`

@@

```{r model_alg, echo = FALSE, warning=FALSE, message=FALSE, cache = TRUE}
n_obs <- nrow(dfa)
threshold <- 0.1
threshold_miss_hrs <- 17

@@

At each sensor location with small proportion of missing values, we estimate a generalised linear model with:
$$\mu_{\text{Sensor}, \text{Month}, \text{DayType}, \text{Time}} \sim \text{Month} + \text{Time}\times \text{DayType}$$
and a quaispoisson error distribution.
- Quasipoisson regression allows for $Var[\text{HourlyCount}] = \theta \mu$ (ie. overdispersion)
- These models only use time and date based variables to predict pedestrian counts
  * `DayType` is a factor level/categorical variable based on the day of the week and public holidays
  * `Time` is treated as a factor level, as it has a non-linear relationship with counts

@@

```
- Need to use data from neighbouring sensors
- Neighbour defined as geographical neighbours for simplicity
  * Find two closest sensors by great-circle (haversine) distance

@@

---
class: middle
## _Step 2c: Fit models using neighbours_
At each sensor location with "large" proportion of missing values, we estimate a generalised linear model with:

@@

---
## Step 3: Replacing missing values with imputed values

Using the models estimated, we can impute the missing values
- For sensors with large proportion of missing values, we needed to have complete data at the neighbours
  * Can't use neighbours with large proportion of missing values
  * Use imputed data from sensors with small proportion of missing values

```{r, echo = FALSE, fig.retina = 6, message=FALSE, warning=F, fig.width=10, fig.height = 4}
p1 <-   ggplot(original_df %>% filter(Date_Time < "2017-01-01")) +
        geom_line(aes(x = Date_Time, y = `Southbank`)) +
        scale_y_continuous(limits = c(0, 7500)) +
        theme_minimal() + ylab("Pedestrian Count") +
        ggtitle("Southbank Pedestrian Counts before Imputation")

p2 <-   ggplot(original_df %>% filter(Date_Time < "2017-01-01")) +
        geom_line(aes(x = Date_Time, y = dfa2$Southbank),
                  colour = viridis(3)[1]) +
        geom_line(aes(x = Date_Time, y = `Southbank`)) +
        scale_y_continuous(limits = c(0, 7500)) +
        theme_minimal() + ylab("Pedestrian Count") +
        ggtitle("Southbank Pedestrian Counts after Imputation")
grid.arrange(p1, p2, ncol = 1)
```

---
class: middle
# Implementation of the imputation algorithm

- general code which will run the algorithm
- inputs required: 
  * data in wide format
  * threshold value for missing values proportion (_default:_ `threshold = 0.1`)
  * max length of repeated values (_default:_ `maxlength = 17`)
- use R and RStudio with packages:
  * `dplyr`, `lubridate` and `tidyr` for data manipulation
  * `foreach` and `doSNOW` for parallel computing
  * `rwalkr` for accessing latest data
---
## Imputation Model Evaluation

```{r, echo = FALSE, fig.retina = 6, message=FALSE, warning=F, fig.width = 9}
p1 <-   ggplot(data = NULL, aes(x = original_df %>% 
             filter(Date_Time < "2017-01-01") %>%
             .[, "Southern Cross Station"],
           y = fitted_data$`Southern Cross Station`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.5, size = 1.5,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.5, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + theme(aspect.ratio = 1) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count") +
           ggtitle("Actual against Predicted at Southern Cross Station")

p2 <-   ggplot(data = NULL, aes(x = dfa %>% 
             filter(Date_Time < "2017-01-01") %>%
             .[, "Southbank"],
           y = fitted_data$`Southbank`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.5, size = 1.5,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.5, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + theme(aspect.ratio = 1) +
           scale_x_continuous(limits = c(0, 6000)) +
           scale_y_continuous(limits = c(0, 6000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count") +
           ggtitle("Actual against Predicted at Southbank")
grid.arrange(p1, p2, nrow = 1)
```
